In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-18 trunk/tex/sigmasfr/ms.tex
wrote section 3.1

\altaffiltext{5}{Institut de Ci{\'e}ncies del Cosmos, University of
  Barcelona, Barcelona, Spain}

@@

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $1000$~\kms\ outflows.  The implied
star formation rate (SFR) surface densities approach
$\sigmasfr=5000$~\msun~yr$^{-1}$~kpc$^{-2}$, and we argue that
feedback associated with this starburst event is a likely mechanism
for launching the high-velocity outflows.

@@

\subsection{HST/WFC3 Imaging}

@@

\subsection{Photometry, Stellar Masses, and Star-formation rates}

We gathered photometry from the All-Sky Release of the Wide-field
Infrared Survey Explorer \citep[WISE,][]{wri10}, the Seventh Data
Release of the Sloan Digital Sky Survey \citep[SDSS,][]{aba09}, and
General Release 6 from the Galaxy Evolution Explorer (GALEX).  We also
obtained $5\times30$~sec dithered exposures in the 3.6~$\mu$m and
4.5~$\mu$m for all sources with the Infrared Array Camera
\citep{faz04} on the Spitzer Space Telescope \citep{wer04} as part of
General Observer program 60145 (PI: C. Tremonti).  We used the
post--basic calibrated data to perform aperture photometry on all
sources, and we also used the
APEX\footnote{http://irsa.ipac.caltech.edu/data/SPITZER/docs/dataanalysistools/tools/mopex/}
point-source extraction software for photometry of sources in crowded
fields.  All the photometry was corrected for Galactic extinction
based on the \citet{sch98} dust maps.  We show spectral energy
distributions (SEDs) for the three high-$\sigmasfr$ galaxies in
Figure~\ref{fig:seds}.

We use the method of \citet{mou11} to derive stellar masses and
UV-based SFRs from fits to the 0.1--5~$\mu$m SEDs.  We also estimate
IR-based SFRs from the 12~$\mu$m and 22~$\mu$m WISE photometry.

Comment on agreement between SFR indicators.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{spectra.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  These spectra are
  dominated by light from the young stellar population and exhibit
  \oii~$\lambda3727$, H$\beta$~$\lambda4861$, and \oiii~$\lambda5007$
  emission lines.  The panels on the right zoom on the
  \mgii~$\lambda\lambda2796,2803$ spectral region in both wavelength
  and velocity space.  The spectrum in the bottom panel has sufficient
  spectral resolution ($\textnormal{FWHM}\approx8$~\kms) to resolve
  the intrinsic shape of the absorption-line profile, revealing that
  the gas near the centroid velocity has unity covering factor.}
\label{fig:spectra}
\end{figure}

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} %\\
%\includegraphics[angle=0,scale=.7]{sizemass_sigmasfr.ps}
%\caption{$\sigmasfr$--mass and size--mass relationships.}
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%
  contours).  We mark the empirical threshold for launching winds
  \citep[dotted line, $\sigmasfr=0.1$~\units;][]{hec02}, the
  90th-percentile starburst intensity limit from \citet{meu97} (dashed
  line, $\sigmasfr=45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit.  The sample of gas-rich mergers includes ULIRGs
  \citep{vei06}, Lyman break analogs with dominant central objects
  \citep{ove09}, and the local compact starburst Arp 220 \citep{sco97,
    ken98, rod08}.  The samples of star-forming galaxies are drawn
  from \citet{wuy11} at $0.02<z<0.2$ (solid black contour),
  $0.5<z<1.5$ (dashed red countour), and $1.5<z<2.5$ (dotted purple
  contour), and from \citet{law12} at $2.5<z<3.6$ (dash-dotted blue
  contour). }
\label{fig:sigmasfr}
\end{center}
\end{figure*}

\subsection{Optical Spectroscopy}

We present low-resolution $R=1000-3000$ spectroscopy from MMT/Blue
Channel and SDSS (J1506+5402), Magellan/MagE (J1341-0321), and
Keck/LRIS (J0905+5759) in Figure~\ref{fig:spectra.}.  We also
highlight the \mgii~$\lambda\lambda2796,2803$ absorption lines, which
are used use to measure outflow velocities.

For J0905+5759 we show the \mgii\ absorption-line profiles from
$R\approx37,000$ spectra obtained with Keck/HIRES.  The oscillator
strength of \mgii~$\lambda$2796 is twice that of \mgii~$\lambda$2803,
but the equivalent widths of the two lines and the shapes of the line
profiles are quite similar, indicating that both lines are optically
thick.  The fact that the absorption line profile approaches zero near
the velocity centroid $v=-2470$~\kms\ indicates that the outflowing
gas covers the entire continuum source at these velocities.

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx500$~\msun) for the galaxies shown in
Figure~\ref{fig:seds} imply extremely large star-formation rate
surface densities ($\sigmasfr\approx5000$~\units).  To place these
galaxies in context, we plot their $\sigmasfr$ values as a function of
stellar mass in Figure~\ref{fig:sigmasfr}.  We also include
measurements of late-stage, gas-rich mergers at $z<0.3$ from
\citet{vei06}, star-forming galaxies at $z<0.3$ from \citet{ove09},
and star-forming galaxies at $z=1.5$--3.6 from \citet{law12}.

Points to make:
\begin{itemize}
\item \citet{ken98}, the standard reference for the relationship
  between SFR and gas surface density, includes Arp 200 with
  $\sigmasfr\approx10^3$~\units\ and
  $\Sigma_{H_2}\approx5\times10^{10}$~\msun~kpc$^{-2}$.  
\item The star clusters considered by \citet{meu97} extend into this
  $\sigmasfr>10^{3}$~\units\ range.

\end{itemize}

@@

@@

Question of whether other observables are plausible.  
\begin{itemize}
\item Expected velocity dispersion.  Some results have suggested that
  turbulence becomes large at high surface densities, and therefore
  the velocity dispersion should be large.  
\item The dynamical time for the starburst.  If gas is consumed with
  an efficiency of a few \% per dynamical time, how long do we expect
  the starburst to last?
\end{itemize}

Other points: 
\begin{itemize}
\item Is this mechanism viable for the whole sample?  We see large
  outflow velocities in rest of the sample, where the galaxies are
  more ``post'' starburst.  This fits in a scenario where the genuine
  post-starbursts are further along in the evolution after the outflow
  quenches star formation.
\item Will these evolve into ellipticals, or do they have enough gas
  to reform a disk?  If they do quench, how difficult would it be to
  grow in size to match the local size--mass relationship?  Could
  their descendants be the handful of local galaxies that don't lie on
  the size--mass relationship?
\item Is there disagreement between $L_{IR}+L_{UV}$ and H$\alpha$
  SFRs?  \citet{hec11} pointed out that this could be a sign of
  escaping ionizing photons.
\end{itemize}

\subsection{Limits on Ongoing AGN Activity}

No evidence in spectra for unobscured AGN component.  J1506+5402 has
X-ray and \nev.

@@

They are rare and thus represent an unusual or short-lived phase.

@@

Connection to ellipticals.  Conceivable that they could grown in size
sufficiently by $z=0$?

@@

@@

@@

@@

>>>>

\altaffiltext{5}{Institut de Ci{\'e}ncies del Cosmos, University of
  Barcelona, 08028 Barcelona, Spain}

@@

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  The
implied star formation rate (SFR) surface densities approach
$\sigmasfr=5000$~\msun~yr$^{-1}$~kpc$^{-2}$, and we argue that
feedback associated with this starburst event is a likely mechanism
for launching the high-velocity outflows.

@@

%\subsection{HST/WFC3 Imaging}
\subsection{Morphologies and Sizes}

@@

%\subsection{Photometry, Star Formation Rates, and Stellar Masses}
\subsection{Star Formation Rates and Stellar Masses}

We gathered photometry from the All-Sky Release of the Wide-field
Infrared Survey Explorer \citep[WISE,][]{wri10}, the Seventh Data
Release of the Sloan Digital Sky Survey \citep[SDSS,][]{aba09}, and
General Release 6 from the Galaxy Evolution Explorer
\citep[GALEX,][]{mar05galex}.  We also obtained $5\times30$~sec
dithered exposures at 3.6~$\mu$m and 4.5~$\mu$m for all sources with
the Infrared Array Camera \citep{faz04} on the Spitzer Space Telescope
\citep{wer04} as part of General Observer program 60145 (PI:
C. Tremonti).  We used the post--basic calibrated data to perform
aperture photometry on all sources, and we also used the
APEX\footnote{http://irsa.ipac.caltech.edu/data/SPITZER/docs/dataanalysistools/tools/mopex/}
point-source extraction software for photometry of sources in crowded
fields.  All the photometry was corrected for Galactic extinction
based on the \citet{sch98} dust maps.  We show spectral energy
distributions (SEDs) for the three high-$\sigmasfr$ galaxies in
Figure~\ref{fig:seds}.

We estimate IR-based star-formation rates for the 25/29 galaxies with
WISE 12 or 22~$\mu$m detections by fitting \citet{cha01} templates to
their 12 and 22~$\mu$m fluxes.  For the 14/29 galaxies with 22~$\mu$m
detections, this yields SFRs that agree with those obtained from the
\citet{ruj12} method based on 24~$\mu$m luminosity with a scatter of
0.05~dex.  We note that several authors have shown that the shape of
the IR SED for star-forming galaxies depends on the surface density of
star formation \citep[e.g.,][]{ruj11,elb11}, with more compact
starbursts having larger total-IR (8--1000~$\mu$m) to mid-IR
(8--24~$\mu$m) ratios.  The extreme $\sigmasfr$ values of our sample
imply large total-IR to mid-IR ratios, characteristic of the most
luminous galaxies in the local universe
\citep[e.g.,][]{cha01,dal02,rie09}.  If we used the most luminous
local templates for the 8/29 sources with SFRs in the ULIRG regime
($\sfrir>200$~\msun~yr$^{-1}$), we would obtain SFRs that are larger
by 0.5~dex than the values we adopt for this paper.

We also estimate star-formation rates and stellar masses based on
stellar population fits to the $\lrest=0.1$--3~$\mu$m SEDs using the
method of \citet{mou11}.  For galaxies with
$\sfrir>50$~\msun~yr$^{-1}$, there is agreement between these UV-based
SFR estimates and $\sfrir$ with a scatter of 0.32~dex.  For an SMC
dust law, we find a median attenuation of $A_V=0.4$~mag.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{spectra.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  These spectra are
  dominated by light from the young stellar population and exhibit
  \oii~$\lambda3727$, H$\beta$~$\lambda4861$, and \oiii~$\lambda5007$
  emission lines.  The panels on the right highlight the
  \mgii~$\lambda\lambda2796,2803$ spectral region in both wavelength
  and velocity space to illustrate the outflow kinematics.  The
  spectrum in the bottom panel has sufficient spectral resolution
  ($\textnormal{FWHM}\approx8$~\kms) to resolve the intrinsic shape of
  the absorption-line profile, revealing that the gas near the
  centroid velocity ($v=-2470$~\kms) covers the entire galaxy.}
\label{fig:spectra}
\end{figure}

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} %\\
%\includegraphics[angle=0,scale=.7]{sizemass_sigmasfr.ps}
%\caption{$\sigmasfr$--mass and size--mass relationships.}
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%,
  95\%, and 99.7\% contours).  We mark the empirical threshold for
  launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit, suggesting growth that is limited by momentum
  injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}

% The sample of gas-rich mergers includes ULIRGs
%   \citep{vei06}, Lyman break analogs with dominant central objects
%   \citep{ove09}, and the local compact starburst Arp 220 \citep{sco97,
%     ken98, rod08}.  The samples of star-forming galaxies are drawn
%   from \citet{wuy11} at $0.02<z<0.2$ (solid black contour),
%   $0.5<z<1.5$ (dashed red countour), and $1.5<z<2.5$ (dotted purple
%   contour), and from \citet{law12} at $2.5<z<3.6$ (dash-dotted blue
%   contour).

%\subsection{Optical Spectroscopy}
\subsection{Outflow Kinematics and Covering Factors}

We present $\lrest=2500$--5200~\AA\ spectroscopy for three
high-$\sigmasfr$ sources in Figure~\ref{fig:spectra} based on data
from MMT/Blue Channel and SDSS (J1506+5402), Magellan/MAGE
(J1341-0321), and Keck/LRIS and Keck/HIRES (J0905+5759).  The spectra
are dominated by light from the young ($t<50$~Myr) stellar population.
We highlight the \mgii~$\lambda\lambda2796,2803$ absorption lines,
which are used used to measure outflow velocities.  At low spectral
resolution (e.g., the top right panel of Figure~\ref{fig:spectra}) it
is not possible to determine the intrinsic shape of the absorption
line profile and therefore the covering factor of the outflowing gas.
However, the Keck/HIRES spectrum of J0905+5759
($\textnormal{FWHM}\approx8$~\kms) reveals that the gas covers the
entire continuum source near the velocity centroid ($v=-2470$~\kms)
indicating a galaxy-wide outflow.

% MMT/Blue Channel (J1506+5402, $R\sim1000$), SDSS (J1506+5402,
% $R\sim2000$), Magellan/MAGE (J1341-0321, $R\sim4000$), Keck/LRIS
% (J0905+5759, $R\sim600$), and Keck/HIRES (J0905+5759, $R\sim37,000$)

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx500$~\msun) for the galaxies described above
implies extremely large star-formation rate surface densities
($\sigmasfr\approx5000$~\units).  To place these galaxies in context,
we plot their $\sigmasfr$ values as a function of stellar mass in
Figure~\ref{fig:sigmasfr}.  We include comparison samples of
$\sim10^5$ star-forming galaxies at $0.5<z<1.5$ from \citet{wuy11} and
a few dozen gas-rich mergers at $z<0.3$ including 32 ULIRGs from
\citet{vei06}, five Lyman break analogs with dominant central objects
from \citet{ove09}, and the local compact starburst Arp 220
\citep{sco97, ken98, rod08}.  We also mark the empirical threshold for
launching winds \citep[$\sigmasfr\approx0.1$~\units,][]{hec02}, the
90th percentile limit for the surface brightness of starbursts
measured using UV, H$\alpha$, far-IR, and radio continuum emission
\citep[$\sigmasfr\approx45$~\units][]{meu97}, and the theoretical
limit for a starburst limited by feedback from radiation pressure
\citep[$\sigmasfr\approx2000$~\units,][]{mur05,tho05,hop10}.  The most
luminous, compact starbursts in our sample exhibit SFR surface
densities that reach the Eddington limit, suggesting that their growth
is being regulated by momentum input from massive stars.

%  We also include measurements of
% late-stage, gas-rich mergers at $z<0.3$ from \citet{vei06},
% star-forming galaxies at $z<0.3$ from \citet{ove09}, and star-forming
% galaxies at $z=1.5$--3.6 from \citet{law12}.

% Points to make:
% \begin{itemize}
% \item \citet{ken98}, the standard reference for the relationship
%   between SFR and gas surface density, includes Arp 200 with
%   $\sigmasfr\approx10^3$~\units\ and
%   $\Sigma_{H_2}\approx5\times10^{10}$~\msun~kpc$^{-2}$.  
% \item The star clusters considered by \citet{meu97} extend into this
%   $\sigmasfr>10^{3}$~\units\ range.
% 
% \end{itemize}

@@

%\section{Discussion}

\subsection{Constraints on Ongoing AGN Activity}

While the SEDs (see Figure~\ref{fig:seds}) and optical spectra (see
Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the amount of ongoing AGN activity in these systems.  The
strongest case for detectable AGN emission among the high-$\sigmasfr$
galaxies can be made for J1506+5402, which has a weak
\nev~$\lambda3426$ emission line typically associated with AGN
activity \citep[e.g., ][]{abe08}.  This galaxy was observed with the
Chandra X-ray Observatory (proposal ID 11700896, PI: C. Tremonti) and
had four detected counts, corresponding to a 0.5--8.0~keV X-ray
luminosity of $10^{42.7}$~erg~s$^{-1}$ (see P. Sell et al., in
preparation for more details on the 12/29 galaxies with Chandra
observations).  Using the relationship between 2--10~keV X-ray and
12.3~$\mu$m mid-IR luminosity for AGNs from \citet{gan09}, one would
expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than observed luminosity for J1506+5402,
$L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$.
If this source were Compton-thick \citep[e.g., the typical 2--10~keV
  attenuation for local Compton-thick AGNs is a factor of
  100;][]{dia09}, then the expected mid-IR AGN contribution could be
relevant at the $\lesssim$30\% level, but this would not affect our
conclusions that this galaxy is starburst dominated with a $\sigmasfr$
value near the Eddington limit.  As is the case for typical ULIRGs
\citep[e.g.,][]{far07}, there is certainly evidence for AGN activity
in a fraction of our sample, but the bolometric output is dominated by
star formation.

% No evidence in spectra for unobscured AGN component.  J1506+5402 has
% X-ray and \nev.  Agreement between SFR indicators.  Mention Balmer
% lines and dust geometry.  SEDs well fit by star-forming tempates.
% Most ULIRGs are starburst dominated.  Mention Pauls's paper.  There is
% likely some ongoing AGN activity in sources like J1506+5402, but in
% general, the AGN does not seem to be relevant energetically.  Point
% that high-velocity outflows don't mean that an outflow is AGN driven.

\subsection{The Outflow Launching Mechanism}

@@

% Question of whether other observables are plausible.  
% \begin{itemize}
% \item Expected velocity dispersion.  Some results have suggested that
%   turbulence becomes large at high surface densities, and therefore
%   the velocity dispersion should be large.  
% \item The dynamical time for the starburst.  If gas is consumed with
%   an efficiency of a few \% per dynamical time, how long do we expect
%   the starburst to last?
% \end{itemize}
% 
% Other points: 
% \begin{itemize}
% \item Is this mechanism viable for the whole sample?  We see large
%   outflow velocities in rest of the sample, where the galaxies are
%   more ``post'' starburst.  This fits in a scenario where the genuine
%   post-starbursts are further along in the evolution after the outflow
%   quenches star formation.
% \item Will these evolve into ellipticals, or do they have enough gas
%   to reform a disk?  If they do quench, how difficult would it be to
%   grow in size to match the local size--mass relationship?  Could
%   their descendants be the handful of local galaxies that don't lie on
%   the size--mass relationship?
% \item Is there disagreement between $L_{IR}+L_{UV}$ and H$\alpha$
%   SFRs?  \citet{hec11} pointed out that this could be a sign of
%   escaping ionizing photons.
% \end{itemize}

@@

They are rare and thus represent an unusual or short-lived phase.  Gas
consumption time for starbursts.  Connection to ellipticals.
Conceivable that they could grown in size sufficiently by $z=0$?

@@

\section{Conclusions}

@@

\bibitem[Abel \& Satyapal(2008)]{abe08} Abel, N.~P., \& Satyapal,
  S.\ 2008, \apj, 678, 686

@@

\bibitem[Diamond-Stanic et al.(2009)]{dia09} Diamond-Stanic, A.~M.,
  Rieke, G.~H., \& Rigby, J.~R.\ 2009, \apj, 698, 623

@@

\bibitem[Elbaz et al.(2011)]{elb11} Elbaz, D., Dickinson, M., Hwang,
  H.~S., et al.\ 2011, \aap, 533, A119
\bibitem[Farrah et al.(2007)]{far07} Farrah, D., Bernard-Salas, J.,
  Spoon, H.~W.~W., et al.\ 2007, \apj, 667, 149

@@

\bibitem[Gandhi et al.(2009)]{gan09} Gandhi, P., Horst, H., Smette,
  A., et al.\ 2009, \aap, 502, 457
